Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-19 manuscript.tex
cleaning up .tex file
edits to discussion

Given that the \emph{Hopscotch} insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the \emph{Hopscotch} in teosinte. While our findings are consistent with \citet{Studer2011} in that we identified the \emph{Hopscotch} allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Many of our populations with high frequency of the \emph{Hopscotch} allele fall in the Jalisco cluster identified by \citet{Fukunaga2005}, possibly suggesting a different history of the \emph{tb1} locus than in the Balsas region where maize was domesticated \citep{Matsuoka2002}. While gene flow from crops into their wild relatives is well-known, \citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}, our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte \citep{Hufford2013}. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize) \citep{Hufford2013}.

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte.  Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1, and assignment to maize in this region in our STRUCTURE analysis is lower than the average across chromosome 1 (Fig. \ref{Fig3Structure}, Table \ref{Table4Q}). Together, these data point to an explanation other than recent introgression for the high observed frequency of \emph{Hopscotch} in some of  our \emph{parviglumis} populations.

@@

Additional explanations for differential frequencies of the \emph{Hopscotch} among teosinte populations include both drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D we present here are consistent with these findings. \jri{deviations from HWE may be consistent too if we see excess of homozygotes. do we?}. This bottleneck, however, does not explain differences in \emph{Hopscotch} allele frequency among multiple populations in the Jalisco group, and the available information on diversity and population structure among these populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of recent colonization or other demographic events that might predict a high frequency of the allele across populations.  \jri{here we need a few sentences on selection.  the 5' UTR has much more negative D than the upstream. do we know the Hop genotype for sequenced lines? can we separate the sequences into hop/no hop and look for differences? it wasn't until we did this that gt1 stuff really popped out.} \lev{we should know for some of them, i will check}
%   from expected genotype frequencies under Hardy-Weinberg equilibrium at the \emph{Hopscotch} insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus..

%\mbh{many populations in the Jalisco cluster of Fukunaga seem to have high frequency of the \emph{Hopscotch} allele.  Why?  Only ancient introgression in Jalisco and not in the Balsas? If \emph{Hopscotch} came from standing variation, is this more consistent with a Jalisco domestication?  I know, heresy!!  LV, can you look at the \emph{parviglumis} populations with high \emph{Hopscotch} and see if they are all in the Jalisco cluster of Fukunaga?  Might be a nice observation we could make here...could even go so far as saying the results from Matsuoka deserve another look.} \lev{Matt, yes they are in the Jalisco cluster, I added in a sentence mentioning it, but am not sure if that's the direction you were wanting me to take with it or not}

>>>>

Given that the \emph{Hopscotch} insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the \emph{Hopscotch} in teosinte. While our findings are consistent with \citet{Studer2011} in that we identified the \emph{Hopscotch} allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Many of our populations with high frequency of the \emph{Hopscotch} allele fall in the Jalisco cluster identified by \citet{Fukunaga2005}, possibly suggesting a different history of the \emph{tb1} locus in this region than in the Balsas River Basin where maize was domesticated \citep{Matsuoka2002}. Potential explanations for the high frequency of the \emph{Hopscotch} element in \emph{parviglumis} from Jalisco include gene flow from maize, genetic drift, and natural selection.

While gene flow from crops into their wild relatives is well-known, \citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}, our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte \citep{Hufford2013}. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize; \citealt{Hufford2013}). We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte.  Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1, and assignment to maize in this region in our STRUCTURE analysis is lower than the average across chromosome 1 (Fig. \ref{Fig3Structure}, Table \ref{Table4Q}). Together, these data point to an explanation other than recent introgression for the high observed frequency of \emph{Hopscotch} in some of  our \emph{parviglumis} populations.

@@

Remaining explanations for differential frequencies of the \emph{Hopscotch} among teosinte populations include both drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D we present here are consistent with these findings. \jri{deviations from HWE may be consistent too if we see excess of homozygotes. do we?}. This bottleneck, however, does not explain differences in \emph{Hopscotch} allele frequency among multiple populations in the Jalisco group, and the available information on diversity and population structure among these populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of recent colonization or other demographic events that might predict a high frequency of the allele across populations.  \jri{here we need a few sentences on selection.  the 5' UTR has much more negative D than the upstream. do we know the Hop genotype for sequenced lines? can we separate the sequences into hop/no hop and look for differences? it wasn't until we did this that gt1 stuff really popped out.} \lev{we should know for some of them, i will check}
%   from expected genotype frequencies under Hardy-Weinberg equilibrium at the \emph{Hopscotch} insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus..
